---
title: "R Markdown Tutorial"
output: 
  rmdformats::robobook:
    #self_contained: true
    thumbnails: false
    lightbox: true
---

This is an introduction to data visualization (in R).


We will be using `ggplot2` to plot our data. Let's make sure you have that installed. 

```{r echo = FALSE}
if (!require("ggplot2")) install.packages("ggplot2")
if (!require("tidyr")) install.packages("tidyr")
if (!require("forcats")) install.packages("forcats")
library(ggplot2)
library(tidyr)
library(forcats)
```

We will use the data from the Baker 2016 paper, again.

```{r}
# Load in our data:
baker2016 = read.table('baker2016_artificial.txt', header = T)
```

```{r, echo=FALSE}
knitr::kable(
  head(baker2016[,c(26,9,16,18)]), # This is the table we will plot
  booktabs = TRUE) # We determine the formatting style as 'booktabs'
```

# Basic Plot Types

The basic structure of a gpplot looks as follows. You call `ggplot`, pass over some `data`, specify aesthetics using `aes()` and detmine the geometry using one (or several) of the following (or many more) options:

- `geom_point()`
- `geom_bar()`
- `geom_boxplot()`
- `geom_col()`
- `geom_hist()`
- `geom_density()`

```{r eval=F, echo=F, message=FALSE, warning=FALSE}
ggplot(data, aes(x = x)) + 
  geom_bar()
```


Let's create our first plot of the data. We are plotting a barplot of the data's column `familiar_reproducibility` which (non surprisingly) shows us how familiary respondents were with the term 'Open Science'.

```{r message=FALSE, warning=FALSE}
ggplot(baker2016, aes(x= flagged)) + 
  geom_bar()
```

Now imagine we wanted to use that same plot structure to plot some other variable, i.e the respondents research field. To do so, we'd just replace `x=familiar_reproducibility` with `x=field`. Easy. Remember, you can check what other columns are available by running `colnames(baker2016)` on the command line. Try out a few. 


Let's try out a different plot type. Histograms can be a great way to plot data distributions. Let's plot participants' responses to the question *'What percentage of the research in your field do you think is reproducible?'*. The column you are looking for is called `proportion_reproducible`, so we would plot the histogram as follows:

```{r message=FALSE, warning=FALSE}
ggplot(baker2016, aes(x = proportion_reproducible)) + 
  geom_histogram(stat="count")
```

# What is your message?

The participants in the study came from different fields. We may be interested in seeing how the responses differed between the fields. GGplot allows you to use color and grouping options in plots. I.e. we may create a stacked histogramm by telling ggplot to `fill` the bars by `field`, specifying `fill=field`. In other plots you may choose the option `color=field` - *color* usually refers to the outlines of plotted items, whereas *fill* well... "fills" them. Feel free to five this a try on the histogramm to see the difference.

```{r message=FALSE, warning=FALSE}
ggplot(baker2016, aes(group= field, fill=field, x = proportion_reproducible)) + 
  geom_histogram(stat='count')
```

Remember we wanted to get an idea of how responses differed between the fields. What are we learning from the previous plot? Probably, not much. So while a histogramm may have helped us understand that most participants appear to have a great amount of trust in the results in their fields, a stacked histogramm might not be the plot that allows us to understand how results differ between fields. 

A good way to disect differences between groups in your data is to use `facet_wrap` to create different panels for each group. Simply add `facet_wrap(facets = baker2016$field)` to get one histogram per group. Hopefully, this helps us see how the distributions differ between the fields. 


```{r message=FALSE, warning=FALSE}
ggplot(baker2016, aes(group= field, fill=field, x = proportion_reproducible)) + 
  geom_histogram(stat = 'density') + 
  facet_wrap(facets = baker2016$field)
```

## Fitting data

The original Baker 2016 data unfortunately did not include continuous variables. I have created artificial data for us to practice plotting continous variables. `dummy_dat2` is an artificial response, and `dummy_age` is an age vector I created based on the age bins from the original data. 

Let's plot the effects of `dummy_age` on `dummy_dat2`.

```{r message=FALSE, warning=FALSE}
ggplot(baker2016, aes(x=dummy_age, y=dummy_dat2)) + 
  geom_point()
```

That looks like a decent correlation. Let's fit a linear model to this, regressing `dummy_age` on `dummy_dat2`. We can fit models to our data with `geom_smooth()`. Model options are linear (`lm`), GLM (`glm`), smoothing splines (`gam`), and loess regression (`loess`). We are using a simple linear model.  


```{r message=FALSE, warning=FALSE}
ggplot(baker2016, aes(x=dummy_age, y=dummy_dat2, color=as.factor(dummy_group))) + 
  geom_point()+
  geom_smooth(method = 'lm')
```

Lastly, I have also created an artificial variable `dummy_group`. You can analyse group effects on your model by simply specifying the color option with your grouping variable. This will then fit two separate linear regressions. In this case we learn that the grouping variable does not appear to have an effect on our data:


```{r message=FALSE, warning=FALSE}
ggplot(baker2016, aes(x=dummy_age, y=dummy_dat2, color=as.factor(dummy_group))) + 
  geom_point()+
  geom_smooth(method = 'lm')
```


# Statistics in figures - or how not to fool your audience (and yourself)

Choosing how to plot your data is crucial to communicate your message. As a first step though you may often use plots to get an idea of what your data looks like. In both cases, make sure to choose a representation that enables you to understand your data well. 

Here, I am creating an artificial dataset to illustrate some points on how (not to) plot your data.

```{r message=FALSE, warning=FALSE}
response1 = rnorm(n = 1000, mean = 10, sd = c(1,2))
response2 = rnorm(n = 1000, mean = c(2,3), sd = 1)
response3 = rnorm(n = 1000, mean = c(4,5), sd = c(1,1.5))
group = rep(letters[1:2], length.out = 100)
df = data.frame(group = as.factor(group), response1=response1, response2=response2, response3=response3)
```

Let's have a look at `response1`. You may think it's a good idea to first plot this data as a barplot, displaying the mean value for `response1` in groups `a` and `b`. This is what such a plot would look like:

```{r message=FALSE, warning=FALSE}
ggplot(df, aes(x=group,y=response1, color=group, fill=group, alpha=0.2)) + 
  geom_bar(stat='summary', fun=mean) 
```

Now that's not very helpful. It looks as if both groups are the same. 

Now you may have noticed that when I created the data, I intentionally created both groups to have the same mean value. However, their standard deviations are very different!

The following plots display the exact same data - but now in a more useful format. 

```{r, figures-side, fig.show="hold", out.width="50%", message=FALSE, warning=FALSE}
par(mfrow=c(2,1))
ggplot(df, aes(x=group,y=response1, color=group, fill=group, alpha=0.2)) + 
  geom_violin()

ggplot(df, aes(x=group,y=response1, color=group, fill=group, alpha=0.2)) + 
  geom_boxplot()
```

Consider having a play around with `response2` and `response3` to see what kind of plots hide or highlight the differences between those variables. 


# Wide and Long Format

The data as it is right now is in the so-called *wide format*. For many types of plots that maynot be useful. 

Data in the *wide format* displays repeated responses in a single row, and each response is in a separate column. Let's have a look at the question: *'Have you ever tried to replicate a result, either your own or someone elses, and failed?'*. Subjects can say yes or no to both the idea that they have tried to replicate someone else's results and failed, coded in `try_failOwn`, and that they have tried to replicate their own work and failed, coded in `try_failElse`. In the *wide format* each subject is a row, and each responses is coded as their own column: 

```{r, echo=FALSE, message=FALSE, warning=FALSE}
knitr::kable(
  head(baker2016[,c(1,20,21,27)]), # This is the table we will plot
  booktabs = TRUE) # We determine the formatting style as 'booktabs'
```

An alternative representation would be to have a column indicating which part of the question subjects repsonded to, i.e. *own* or *someone else's*, with each subject being represented twice: One row encodeing their response to the first, the second row encoding their response to the second question. 

```{r, fig.show="hold", out.width="50%", message=FALSE, warning=FALSE}
tmp = gather(baker2016[,c(1,20,21,27)], question, response, 2:3)
knitr::kable(
  head(tmp[order(tmp$responseid),]), # This is the table we will plot
  booktabs = TRUE)
```

In order to get our data into this format, we make use of the `gather` function from the `tidyr` package. This functions takes in the arguments 

(1) `data`, i.e. the data in wide format to be transformed into long format
(2) `key`, the variable name of a new column which will enumerate the different categories possible
(3) `value`, the variable name of a new column which will contain the actual data
(4) and a selection of columns which will be converted into the entries in your new column. 

```{r, fig.show="hold", out.width="50%",message=FALSE, warning=FALSE}
data_wide = baker2016[,c(20,21,27)] # This is the data we will be using for our plot in wide format 
data_long = gather(data = data_wide, key = question, value = response, 1:2) # Convert the data into long format
```

The data in wide format allows us to plot something like this for example:

```{r message=FALSE, warning=FALSE}
# Delete responses where participants couldn't remember the result
data_long = data_long[-which((data_long$response == "I can't remember")),]

ggplot(data_long[data_long$question=='try_failElse',], aes(x = field, fill=response)) + 
  geom_bar(aes(y = ..count../tapply(..count.., ..x.. ,sum)[..x..]), position = 'dodge',group = 1) +
  scale_fill_manual(values = c("red", "orange"))+
  coord_flip()
```

# Prettier plots
You can change a whole range of aspects about a theme. A comprehensive list of options can be found [here](https://ggplot2.tidyverse.org/reference/theme.html). Here are a few basic things to get started with:


## Color
You can can change the colors of your plot manually, or by using a colormap. In order to manually determine the colors used, add `scale_fill_manual(value = c('color1','color2',...))`. 

R also supplies a range of predefined colormaps you can use. Have a look on page 4 of this [cheatsheet](https://www.nceas.ucsb.edu/sites/default/files/2020-04/colorPaletteCheatsheet.pdf) for a decent overview. 

```{r}
ggplot(data_long[data_long$question=='try_failElse',], aes(x = field, fill=response)) + 
  geom_bar(aes(y = ..count../tapply(..count.., ..x.. ,sum)[..x..]), position = 'dodge',group = 1) +
  scale_fill_manual(values = c("red", "orange"))+
  coord_flip()
```

Instead of manually defining the colors above, we could also have used a predefined discrete color scale, by adding `scale_fill_brewer(palette = 'Your brewer color scale')`.

```{r message=FALSE, warning=FALSE}
ggplot(data_long[data_long$question=='try_failElse',], aes(x = field, fill=response)) + 
  geom_bar(aes(y = ..count../tapply(..count.., ..x.. ,sum)[..x..]), position = 'dodge',group = 1) +
  scale_fill_brewer(palette = 'Set3')+
  coord_flip()
```

Lastly, let's consider using color to highlight. You can add a conditional argument to the `color` or `fill` options to highlight certain aspects of your plot. I.e. we may want to highlight the participants who responded saying they weren't familiar enough with the term *Open Science* - for example in order to convince the department to run additional workshops on Open Science. 

In order to highlight this aspect, we add the conditional argument `ifelse(flagged=='Not enough',1,2)` to our coloroption. Wrapping it in `as.factor()` simply makes is a discrete variable, which would allow us to choose a discrete colormap. 

```{r message=FALSE, warning=FALSE}
ggplot(baker2016, aes(x= flagged, fill = as.factor(ifelse(flagged=='Not enough',1,2)))) + 
  geom_bar()+
  xlab('Response')+
  ylab('Count')+
  ggtitle("How familiar are you with the term 'Open Science'")
```

## Reordering factor
In the previous plots, the order of bars wasn't necessarily helpful at making the graph more easily understandable. Sometimes small things like reordering bars can make the figure more visually appealing and help direct the viewer's attention to your message. 

When factor variables are plotted, the ordering is done by the factor levels. In our case that's the variable `flagged`. You can check what your factor levels are using `levels(baker2016$flagged)`. Currently the order is `"A reasonable amount","I am unsure","Not enough","Too much"`. 

You could choose to manually reorder your factor levels, or simply make use of the `fct_infreq()` function from the `forcats` package, which will sort your factor levels by frequency. 

```{r message=FALSE, warning=FALSE}
ggplot(baker2016, aes(x= fct_infreq(flagged), fill = as.factor(ifelse(flagged=='Not enough',1,2)))) + 
  geom_bar()
```


## Point shapes

You can edit the shapes of your scatter points using a shape from this [table](http://www.sthda.com/english/wiki/r-plot-pch-symbols-the-different-point-shapes-available-in-r) by adding `shape` to `geom_point()`. 

```{r message=FALSE, warning=FALSE}
ggplot(baker2016, aes(x=dummy_age, y=dummy_dat2)) + 
  geom_point(shape=8)+
  geom_smooth(method = 'lm')
```

## Alpha
A super simple way to make your plots look softer can be to add transparency to them. Going back to our scatter plot from earlier, this may actually help to direct attention to important feature of the graph like your regression line and make overlapping data points more distinguishable. Simply add the argument `alpha`.

```{r message=FALSE, warning=FALSE}
ggplot(baker2016, aes(x=dummy_age, y=dummy_dat2, alpha = 0.2, color=as.factor(dummy_group))) + 
  geom_point()+
  geom_smooth(method = 'lm')
```

## Axis labels

Add a title using `ggtitle('Your title goes here.')` and change the axis lables using `xlab('X axis title')` and `ylab('X axis title')`.

```{r message=FALSE, warning=FALSE}
ggplot(baker2016, aes(x= fct_infreq(flagged), fill = as.factor(ifelse(flagged=='Not enough',1,2)))) + 
  geom_bar()+
  xlab('Response')+
  ylab('Count')+
  ggtitle("How familiar are you with the term 'Open Science'")
```

## Change the theme:
Ggplot has built in [themes](https://ggplot2.tidyverse.org/reference/ggtheme.html) that you can make use of. These include: `theme_bw()`, `theme_classic()`, and `theme_void()`. Try out a few and see the effects on this plot:

```{r message=FALSE, warning=FALSE}
ggplot(baker2016, aes(x= fct_infreq(flagged), fill = as.factor(ifelse(flagged=='Not enough',1,2)))) + 
  geom_bar()+
  xlab('Response')+
  ylab('Count')+
  ggtitle("How familiar are you with the term 'Open Science'")+
  theme_bw()
```

## Edit legend

By default, ggplot uses the factor names from your data in the legend, and the respective variable name as the title. 

There are a variety of ways you can edit the legend. Here, we will use `labs()`. 

Since our legend refers to the fill color of the plot, you will have to reference *fill* when editing it. 
You may not want to have a legent title at all, in which case you could add: `labs(fill='')`. Or maybe you would like to rename it to *Questionnaire Response* using `labs(fill='Questionnaire Response')`? Alternatively, you could also decide the legend isn't adding much value here and leave it out entirely by adding `legend.position = "none"`.


```{r message=FALSE, warning=FALSE}
ggplot(baker2016, aes(x= fct_infreq(flagged), fill = as.factor(ifelse(flagged=='Not enough',1,2)))) + 
  geom_bar()+
  xlab('Response')+
  ylab('Count')+
  ggtitle("How familiar are you with the term 'Open Science'")+
  theme_bw() +
  labs(fill='') 
```


## Save plots
Now that we have created lovely plots, we may want to save them as files in order to share and publish them. An easy way to do so is to click on `Export` in the *Plots* panel. 

More conveniently maybe, we can add a bit of code for the plots to automatically autput to a location on our computer. By wrapping your plot in `pdf('/path/to/your/file.pdf', width, height) ... dev.off()`, `png('/path/to/your/file.png', width, height) ... dev.off()`, or `jpeg('/path/to/your/file.jpeg', width, height) ... dev.off()` you can save your plot to different file formats. Note that weirdly `width` and `height` are in *inches* for `pdf()`, whereas for `jpeg()` and `png()` they are in *pixels*. By default the plot should save to your current working directory, so you should change the path to whereever you would love to output the file to. 

```{r message=FALSE, warning=FALSE}
pdf('tmp.pdf',4,5)
ggplot(baker2016, aes(x= fct_infreq(flagged), fill = as.factor(ifelse(flagged=='Not enough',1,2)))) + 
  geom_bar()+
  xlab('Response')+
  ylab('Count')+
  ggtitle("How familiar are you with the term 'Open Science'")+
  theme_bw() +
  theme(legend.position = "none")
dev.off()
```

# Further Resources

This [website](https://www.r-graph-gallery.com/index.html) could be considered the bible of plotting data in R. 
This ggplot [cheatsheet](https://rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf) is great to have at hand. 
